---
title: "RepData_ActivityMonitoring"
output:
  html_document:
    keep_md: yes
---
  

``` {r}
# load the libraries
  suppressPackageStartupMessages(library(data.table))
  suppressPackageStartupMessages(library(dplyr))
  suppressPackageStartupMessages(library(ggplot2))
```

  
### Loading and pre-processing the activity monitoring date.

```{r}
  unzip("activity.zip")
  
  activityData <- read.csv("activity.csv")
  # converting date from factors to date
  activityData$date <- as.Date(activityData$date, "%Y-%m-%d")

```
  
### plot for total number of steps per day
```{r}
  # ignoring the data with NAs
  completeData <- activityData[!is.na(activityData$steps), ]
  
  aggByDate <- aggregate(completeData$steps, by = list(completeData$date), FUN = sum)
  colnames(aggByDate) <- c("date", "totalSteps")
  
  # plot a bar plot not sure why questions refers to a histogram
   g <- ggplot(data = aggByDate, aes(x = date, y = totalSteps))
  g <- g + geom_bar(stat = "identity")
  g <- g + labs(x = "date", y = "total steps")
  g <- g + ggtitle("total number of steps per day")
  print(g)
  
  # mean
  meanSteps <- mean(aggByDate$totalSteps)
  
  # order the data for finding the mean.
  orderedData <- aggByDate[order(aggByDate$totalSteps),]
  medianSteps <- median(orderedData$totalSteps)

```
The mean number of steps taken per day are `r meanSteps`.  
The median number of steps are `r medianSteps`.

  
### Q2 : What is the average daily activity pattern?
```{r}
  aggByInterval <- aggregate(completeData$steps, by = list(completeData$interval), FUN = mean)
  colnames(aggByInterval) <- c("interval", "averageSteps")
  
  plot(aggByInterval$interval, aggByInterval$averageSteps, type = "l", 
       xlab = "interval", ylab = "average_steps",
       main = "average steps in different intervals.")

  # maximum average  number of steps
  maxSteps <- max(aggByInterval$averageSteps)
  maxStepInterval <- aggByInterval[ aggByInterval$averageSteps == maxSteps, "interval"]
```

The interval `r maxStepInterval` has the highest average steps ( `r maxSteps`) across all days.

  
### Imputing missing values
```{r}
# total number of missing values in the dataset.
  missingValues = activityData[is.na(activityData$steps),]
  missingValuesCount = nrow(missingValues)
```
The total number of missing values are `r missingValuesCount`.

```{r}
# filling missing value in any interval by getting the average steps for that interval across all days.
# creating a data.table for faster access.
  avgStepsByInterval <- data.table(aggByInterval)
  setkey(avgStepsByInterval, interval)
  
  # copy the data into a new data set.
  filledActData <- activityData
  
  # impute function
  extrapolatingFunction <- function(actData) {
    dataSteps <- as.numeric(actData["steps"])
    dataInterval <- as.numeric(actData["interval"])
    if (is.na(dataSteps)) {
      avgStepsByInterval[interval == dataInterval]$averageSteps
    }
    else {
      dataSteps
    }
  }

  imputedSteps <- apply(filledActData, 1, extrapolatingFunction)

  # filling the empty values in the new data set.
  filledActData$steps <- imputedSteps

```

``` {r}
  # plot for imputedDataSet
  filledAggByDate <- aggregate(filledActData$steps, by = list(filledActData$date), FUN = sum)
  colnames(filledAggByDate) <- c("date", "totalSteps")
  
  g <- ggplot(data = filledAggByDate, aes(x = date, y = totalSteps))
  g <- g + geom_bar(stat = "identity")
  g <- g + labs(x = "date", y = "total steps")
  g <- g + ggtitle("imputed data - total number of steps per day")
  print(g)
  
  # finding mean
  filledMeanSteps <- mean(filledAggByDate$totalSteps)
  
  # finding median (ordering tthe data for finding the median)
  filledMedianSteps <- median(filledAggByDate[order(filledAggByDate$totalSteps), "totalSteps"])

```
The mean for filled data set is `r filledMeanSteps`.  
The median for filled data set is `r filledMedianSteps`.  
so we see that the mean is same as before, but the median has changed for the filled data set.

  
### Are there differences in activity patterns between weekdays and weekends?

```{r}
  # get the weekday and weekend as a column in the dataset.
  isWeekend <- grepl("^(Saturday|Sunday)$", weekdays(filledActData$date))
  # factor variable indicating weekday and weekend.
  isWeekend <- as.factor(isWeekend)
  levels(isWeekend) <- c("weekday", "weekend")

  filledActData <- mutate(filledActData, isWeekend)

  aggByIntervalAndWeekend <- aggregate(steps ~ interval + isWeekend, data = filledActData, mean)


  g <- ggplot(aggByIntervalAndWeekend, aes(x = interval, y = steps))
  g <- g + geom_line()
  g <- g + facet_grid(isWeekend ~ .)
  g <- g + labs(x = "interval", y = "total steps")
  g <- g + ggtitle("average steps across different intervals/weekend/weekdays")
  print(g)
```

